The Civil Society Organizations effect: A mixed-Methods analysis of bottom-up approaches in Brazilian public policy.

### 2023-11-06 Untitled/Untitled.tex
paper R&P

---
title: "The Civil Society Organizations effect: A mixed-Methods analysis of bottom-up approaches in Brazilian public policy."
short: "The Civil Society Organizations effect"
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
keywords:
  - accountability
  - Civil Society Organizations
  - mixed methods
  - community-driven development
  - public policy.
author:
  - name: Manoel Galdino
    firstname: Manoel
    surname: Galdino
    email: mgaldino@usp.br
    affiliation: Universidade de São Paulo, Brazil
  - name: Bianca Vaz Mondo
    firstname: Bianca
    surname: Mondo
    affiliation: Government Transparency Institute - Germany
  - name: Juliana Mari Sakai
    firstname: Juliana
    surname: Sakai
    affiliation: Transparência Brasil - Brazil
  - name: Natália Paiva
    firstname: Natália
    surname: Paiva
    affiliation: Alandar Consulting - Germany
acknowledgements: |
  We would like to thank all participants of project Obra Transparente, Vanda Medeiros and the team of Transparência Brasil who worked in the project or in evaluations of the project. The project was funded by an UNDEF grant, which we also thank.
abstract: |
  This paper examines the effect of bottom-up accountability on public service delivery. We differentiate between information-driven interventions and the mobilization and monitoring efforts of organized Civil Society Organizations (CSOs), and argue that the latter type of interventions can can drive significant policy change. The study evaluates the effectiveness of the Obra Transparente project by Brazilian NGO Transparência Brasil, engaging 21 local CSOs in South and Southeast Brazil.  Our findings emphasize the importance of sustained, coordinated efforts by socially embedded CSOS. These efforts, involving direct engagement with municipal officials and ensuring that their complains cannot be ignored lead to more substantial outcomes when compared with citizens information-driven.
bibliography: OT_paper.bib
output: rticles::aea_article
---

```{r package loads, message=F, echo=FALSE}
library(data.table)
library(here)
library(janitor)
library(lubridate)
library(readxl)
library(tidyverse)
library(gt)
library(ipeadatar)
library(rstan)
library(arm)
library(rstanarm)
library(survey)
library(broom)
library(broom.mixed)
library(MetBrewer)
```

```{r loading construction data,  message=FALSE, echo=FALSE}

# load(here("obra_transparente", "Dados", "obras_inicio_projeto.Rdata"))
# load(here("obra_transparente", "Dados", "obras_fim_seg_fase.Rdata"))

load(here("Dados", "obras_inicio_projeto.Rdata"))
load(here("Dados", "obras_fim_seg_fase.Rdata"))

# simec23 <- fread(here("obra_transparente", "Dados", "simec 25-10-23 - simec.csv"), encoding = "UTF-8")
simec23 <- fread(here("Dados", "simec 25-10-23 - simec.csv"), encoding = "UTF-8")

# load(here("obra_transparente", "Dados", "obras_fim_seg_fase.Rdata"))

load(here("Dados", "obras_fim_seg_fase.Rdata"))

df_cidades_ot <- read_excel(here("Dados", "situacao obras_atual.xlsx"), 
                                           sheet = "SIMEC-Mai17", skip = 1) %>%
  clean_names() %>%
  filter(!is.na(id)) %>%
  dplyr::select(id, municipio, uf, situacao2, levant_os_jun_17) %>%
  mutate(indicator_ot = 1)
```

```{r loading-processing ipea data,  message=FALSE, echo=FALSE, cache=TRUE}
aux_muni <- geobr::lookup_muni(code_muni = "all") %>%
  dplyr::select(-c(name_state, name_intermediate))

ipea_idhm <- ipeadatar::ipeadata("ADH_IDHM") %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(idhm = value)

ipea_pea18 <- ipeadatar::ipeadata("ADH_PEA18M")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(pea18 = value)

ipea_ext_pobres <- ipeadatar::ipeadata("ADH_PIND")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(ext_pobres = value)

ipea_ext_pobres_neg <- ipeadatar::ipeadata("ADH_PIND_NEG")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(ext_pobres_negros = value)

ipea_rpc <- ipeadatar::ipeadata("ADH_RDPC")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(rpc = value)

ipea_rpc_mulheres <- ipeadatar::ipeadata("ADH_RDPC_MUL")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(rpc_mulheres = value)

ipea_rpc_negros <- ipeadatar::ipeadata("ADH_RDPC_NEG")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(rpc_negros = value)

ipea_ext_prop_pobres <- ipeadatar::ipeadata("ADH_PMPOB")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(ext_prop_pobres = value)

ipea_10p_ricos_desigualdade <- ipeadatar::ipeadata("ADH_PREN10RICOS")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(p10_ricos_desigualdade = value)

ipea_freq_escola_15_17 <- ipeadatar::ipeadata("ADH_T_FREQ15A17")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(freq_escola_15_1 = value)

ipea_analfabetismo <- ipeadatar::ipeadata("ADH_T_ANALF15M")  %>%
  dplyr::filter(date > "2001-01-1" & as.character(uname) == "Municipality") %>%
  rename(analfabetismo = value)

pop2010 <- ribge::habitantes2010 %>%
  mutate(tcode = as.integer(paste0(codigo_uf, codigo_munic)))

ipea_final <- ipea_idhm %>%
  inner_join(aux_muni, by = join_by(tcode == code_muni)) %>%
  inner_join(ipea_pea18, by = "tcode") %>%
  inner_join(ipea_ext_pobres, by = "tcode") %>%
  inner_join(ipea_rpc, by = "tcode") %>%
  inner_join(ipea_ext_prop_pobres, by = "tcode") %>%
  inner_join(ipea_10p_ricos_desigualdade, by = "tcode") %>%
  inner_join(ipea_freq_escola_15_17, by = "tcode") %>%
  inner_join(ipea_analfabetismo, by = "tcode") %>%
  inner_join(pop2010, by = "tcode") %>%
  dplyr::select(-contains("date")) %>%
  dplyr::select(-contains("uname")) %>%
  dplyr::select(-starts_with("code"))
```

```{r processing data, echo=FALSE,  message=FALSE}

obras_inicio_projeto_ot <- obras_inicio_projeto %>%
  left_join(df_cidades_ot, by=join_by("municipio", "uf", "id"))

obras_inicio_projeto_ot <- obras_inicio_projeto_ot %>%
  mutate(concluida = grepl("Concluída", situacao))

aux_indicator_ot <- df_cidades_ot %>%
  distinct(municipio, uf) %>%
  mutate(indicator_muni_ot = 1)

obras_inicio_projeto_ot <- obras_inicio_projeto_ot %>%
  mutate(indicator_ot = ifelse(is.na(indicator_ot), 0, indicator_ot)) %>%
  left_join(aux_indicator_ot, by = join_by("municipio", "uf")) %>%
  mutate(indicator_muni_ot = ifelse(is.na(indicator_muni_ot), 0, indicator_muni_ot))

antes_projeto <- obras_inicio_projeto_ot %>%
  dplyr::filter(uf %in% c("SP", "PR", "SP", "SC", "RS", "MG")) %>%
  group_by(indicator_muni_ot, concluida) %>%
  summarise(contagem = n()) %>%
  mutate(total = sum(contagem),
         perc = round(contagem/total, 2))

antes_projeto <- antes_projeto %>%
  dplyr::filter(concluida == TRUE) %>%
  dplyr::select(-c(concluida, contagem, total)) %>%
  ungroup()

obras_fim_seg_fase <- obras_fim_seg_fase %>%
  mutate(concluida = as.numeric(grepl("Concluída", situacao)),
         id = as.numeric(id)) %>%
  left_join(df_cidades_ot, by=join_by("municipio", "uf", "id")) %>%
  mutate(indicator_ot = ifelse(is.na(indicator_ot), 0, indicator_ot)) %>%
  left_join(aux_indicator_ot, by = join_by("municipio", "uf")) %>%
  mutate(indicator_muni_ot = ifelse(is.na(indicator_muni_ot), 0, indicator_muni_ot))

depois_projeto <- obras_fim_seg_fase %>%
  dplyr::filter(uf %in% c("SP", "PR", "SP", "SC", "RS", "MG")) %>%
  group_by(indicator_muni_ot, concluida) %>%
  summarise(contagem = n()) %>%
  mutate(total = sum(contagem),
         perc = round(contagem/total, 2))

depois_projeto <- depois_projeto %>%
  dplyr::filter(concluida == TRUE) %>%
  dplyr::select(-c(concluida, contagem, total)) %>%
  ungroup()

```

## Introduction
Accountability is a crucial element in democratic systems, ensuring the efficient delivery of public services [@besley_2003; @cameron_2004; @odonnell_1998]. Initially defined as the process of holding authorities responsible for their actions [@mulgan_2000; @oloughlin_1990], the literature has since broadened to include various oversight mechanisms for politicians and the bureaucracy. Some authors focus on formal institutions [@odonnell_2003; @kenney_2003], known as horizontal accountability, while others examine how politicians respond to voters, the media, and civil society organizations [@bertelli_2020; @relly_2012; @paul_1992], referred to as vertical accountability.

Recent studies have highlighted the role of "bottom-up accountability" (also known as social accountability, Fox, 2015), in which citizens receive information about public policies and use it for social control. The pioneering study 'Power to the People' [@bjorkman_2009], which demonstrated the impact of local community engagement in monitoring the provision of health services, inspired subsequent literature that tested similar interventions in different contexts.

The mechanism would works like this: Constituents have firsthand information about the outcomes of local policies, giving citizens incentives to combat corruption that directly affects them. Since policymakers are sensitive to social feedback from their own communities [@serra_2012], bottom-up accountability could influence local public policies

However, a series of recent studies have raised doubts about the effectiveness of "bottom-up accountability" strategies based on interventions aimed at increasing the monitoring capacity of ordinary citizens regarding public policies [@freire_etal2020; @raffler_etal2019; @fox_2015].

@@

The disappointing results have cast doubts on the effectiveness of the bottom-up accountability strategy for improving the delivery of public services, reducing corruption, and enhancing government efficiency. Some authors [@fox_2015; @ankamah_2019] have suggested that social accountability without "teeth," defined as the government's capacity to respond to citizens' voices, has limited impact.

While it has been consistently shown that tactical interventions have, at best, a small, limited impact, it doesn't mean that voice without "teeth" is an ineffective strategy. The argument in this study is that not all types of interventions that promote citizen's voices are the same.

We argue that we need to differentiate between information-led tactical interventions and active mobilization and monitoring by Civil Society Organizations (CSOs). By their very nature, established CSOs are embedded in society and politics and can have a substantial impact on policy change [@gong_2017; @joshi_2012].

The paper argues that a missing feature in most of the literature is that these interventions should target CSOs instead of individual citizens, and ensure that the intervention enhances the ability of the CSOs to overcome their collective action problem [@mattoni_2021; @odilla_2023].

The present paper provides evidence supporting this argument. We reassess the evidence provided by @freire_etal2020, considering a different intervention than the one they evaluated. They provided experimental evidence that a mobile app (Tá de Pé) made available to citizens did not improve the delivery of school and nursery construction works. Here, we assess the efficacy of a sister project, called "Obra Transparente," also developed by the Brazilian NGO Transparência Brasil. This project targeted 21 local Civil Society Organizations in 21 municipalities in the South and Southeast regions of Brazil, with the aim of improving access to education by addressing mismanagement and delays in the construction of public schools and nurseries. Both projects were developed in parallel, with similar goals but achieved strikingly different results.

The argument in the present study is that not all types of interventions that promote citizens' voice are the same. We argue that we need to differentiate between information-led tactical interventions and active mobilization and monitoring by Civil Society Organizations. By their very nature, established CSOs are embedded in society and politics and, as such, can have teeth of their own [@gong_2017; @joshi_2012].  It may be only that we need mobilization by organized civil society, in a structured way, for civil society voices to be heard with “teeth”. Some studies have suggested that a key feature of successful change in public policies is the pressure of collectively organized agents [@mattoni_2021; @odilla_2023]. Collective action is a hard problem to be overcome and creating CSOs is one way to achieve this. We argue in this paper that a missing features of the tactical interventions assessed in most of the literature is that they should target CSOs instead of individual citizens, and make sure that the intervention increases the ability of the CSOs to attract and overcome their collective action problem.

The "Obra Transparente" project consisted of developing a network of local CSOs to share experiences and information on monitoring the construction works, providing a technical support system for monitoring the process of constructing public schools and nurseries, and training on monitoring biddings, contracts, and constructions of public schools and nurseries.

@@

## The project Obra Transparente

@@

## Research Design and Methods

We employ an integrated inference approach [@humphreys_2023] to analyze qualitative data within the framework of causal queries. As this approach is relatively innovative within qualitative research, we'll provide a comprehensive explanation.

The methodology outlined by [@humphreys_2023] mandates the creation of a causal model encoded in Directed Acyclic Graphs (DAGs), a framework developed by Pearl (2009). Bayesian inference, recently promoted in qualitative data analysis [@fairfield_2022], serves as the foundation for deriving quantitative estimates for causal queries constructed using the proposed DAG.

In our current study, we investigate the impact of Civil Society Organization (CSO) monitoring on two aspects: the completion of construction works and the rectification of issues and irregularities found in construction facilities. Formally, the first research question relates to a causal relationship between monitoring (M) and completion (C), indicated by the DAG (M -> C). The second research query delves into whether monitoring (M) causes the fixing of issues (F), or the (DAG M -> F). To accommodate the qualitative nature of our variables, we treat them as binary.

@@

This approach hinges on the notion that each case can belong to a specific type, denoting different causal relationships. In our current investigation, let's focus on the (M -> C) DAG. To clarify what a type is, we'll employ the Potential Outcomes (PO) notation [@rubin_2005; @imbens_2015]. Within the PO framework, every case has two potential outcomes: one when the treatment (in this case, CSO monitoring, M) is absent and another when the treatment is present. In simplified notation, if a city only completes construction works in the absence of monitoring, we express this in PO notation as C(1) = 0. If there is no monitoring (M = 0) and no completion, then we write C(0) = 0. Conversely, if there is monitoring (M = 1) and completion, we write C(1) = 1, and C(0) = 1 if there is completion without monitoring.

We can categorize a city into one of four types based on what happens when they do or do not receive the treatment (monitoring):
Adverse (type "a"): A city that completes construction only if it doesn't receive the treatment (monitoring).
Beneficial (type "b"): A city that completes construction only if there is monitoring.
Chronic (type "c"): Cities that never complete construction, regardless of monitoring.
Destined (type "d"): Cities that always complete construction, regardless of monitoring.
The PO notation is instrumental here, as we can rewrite the types based on the following notation: Type "a" is one in which the potential outcome is 0 if M = 1 and 1 if M = 0. Therefore, we express it as $\theta_{10}$, where the first subscript denotes the outcome when the treatment is 0, and the second denotes the outcome when the treatment is 1. We employ analogous notation for all other types: $b = \theta_{01}$, $c = \theta_{00}$, and $d = \theta_{11}$.

When analyzing a case study of what happened in a given municipality, we can use the data to answer, for example, the following causal query: what is the probability that the city if type “b”, i.e, benefits from the monitoring and would have not completed the construction work if not for the monitoring of the local CSO?

The advantage of this approach is that all assumptions and restrictions imposed to identify the parameters of interest are transparent. It also allows the user to update posterior probabilities based on qualitative data. One reason for this is that with Bayesian inference, we do not need to rely on asymptotic theory to make inferences. Thus, there is no need for a large-n study. Formally speaking, one can make causal queries without any data, based only on prior probabilities for the parameters. Obviously the answer to such causal queries will be vague enough to be almost useless. But the point is that, once you have a prior and a causal model, encoded in a DAG, you can answer causal queries. As one collects data (any data, be it qualitative or quantitative), one can make assumptions about the nature of the data (such as the Data Generating Process), and then update the model and provide new answers to the causal queries, which hopefully will be informative enough to be useful for the task at hand. And even when the model or hypothesis is not completely identified, it can be partially identified [@manski_2003], which is still a useful answer to causal queries.

## Data

@@

## Case Studies

### Delivery of the construction works

@@

To put in perspective what was accomplished regarding the completion of all four projects, by the end of the project, of all 135 construction works planned for all 21 municipalities, only 25 were effectively delivered, a 18.5% delivery rate. In Taubaté,  before the monitoring by the local CSO, the delivery rate was zero. After that, the delivery rate was 100% and one of the construction work was a previously paralyzed construction. According to a report by Transparência Brasil [@coelho_etal2021], out of 771 stalled construction projects that were eventually completed, over 50% of them took more than two years to be delivered and 25% took three years or more. As we can see, the odds were against finishing the construction works.

The local CSO monitoring the construction work in Taubaté did one of  the best jobs among all 21 municipalities covered in the project. In fact, it was mostly the work of a single volunteer with support from the local OSB and Transparência Brazil throughout the project. She monitored four ongoing construction works on a periodic basis, she was very knowledgeable and insistent in demanding response from contractors and the local government. To give a single example of the quality of the social monitoring going on, in an email sent on 17th January of 2018 to the local government, the local OSB volunteer wrote: “Fazendinha Daycare: a. construction log verified. There have been a few more construction workers on the site since December 15, 2017, but not enough, especially considering the extensive amount of work to be completed by the new delivery date in April 2018. b. (...). c. Lack of roof sealing, leading to water entering the classrooms (a situation that already existed on November 14, 2017)...” . She was by far the most knowledgeable and zealous volunteer of the whole project.

@@

From a causal perspective, we are interested in what is called case-level causal attribution [@humphreys_2023], i.e, did the monitoring cause the delivery of the construction works? In other words, we are conditioning on the fact that M = 1 and C = 1, which constrains the possible types for Taubaté to be $\theta^C_{1,1}$ ou $\theta^C_{0,1}$. Assuming a simple model in which CSO monitoring causes the completion of works (M -> C), and with the eight within-case instances, we do the causal query. Our stimand of interest is the probability that Taubaté is a type $\theta^D_{10}$.

Assuming uniform priors, the probability that  Taubaté is a type $\theta^D_{10}$, i.e, it is a city in which the treatment has a positive effect is 65%, with 95% credibility interval around [23%, 95%]. As we can see, a wide interval, due to the low sample size, and yet, a positive effect nonetheless with 95% of certainty.

What are the assumptions behind such a result? The main one is that we modeled the monitoring (the treatment) as-if random, which is obviously not true. The mere passage of time causes an increased probability of delivery of a construction work. We can also imagine other possible confounds. The way to model it in a DAG is to add an arrow from M to C and vice-versa (M <-> C). Now, the estimand of interest depends on the probability of the treatment (monitoring) causing also the outcome (Average Treatment Effect, ATE). We will not detail the derivation of the new stimand here (you can consult annex B), but the idea is similar to before: we are interested in the type of city in which there is a causal effect of the monitoring on the completion of a construction works of school and/or nurseries. The mean probability of Taubaté being of this type, assuming the model is true is 43%, with 95% credibility intervals around [9%, 80%]. The interval is wider because we added the uncertainty of confounding. Nonetheless, it is unlikely that the positive effect found in the observational data is driven only due to the confounding variables.

>>>>

% AER-Article.tex for AEA last revised 22 June 2011
\documentclass[]{AEA}

% The mathtime package uses a Times font instead of Computer Modern.
% Uncomment the line below if you wish to use the mathtime package:
%\usepackage[cmbold]{mathtime}
% Note that miktex, by default, configures the mathtime package to use commercial fonts
% which you may not have. If you would like to use mathtime but you are seeing error
% messages about missing fonts (mtex.pfb, mtsy.pfb, or rmtmi.pfb) then please see
% the technical support document at http://www.aeaweb.org/templates/technical_support.pdf
% for instructions on fixing this problem.

% Note: you may use either harvard or natbib (but not both) to provide a wider
% variety of citation commands than latex supports natively. See below.

% Uncomment the next line to use the natbib package with bibtex
\usepackage{natbib}

% Uncomment the next line to use the harvard package with bibtex
%\usepackage[abbr]{harvard}

% This command determines the leading (vertical space between lines) in draft mode
% with 1.5 corresponding to "double" spacing.
\draftSpacing{1.5}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}

\usepackage{hyperref}

\begin{document}

\title{The Civil Society Organizations effect: A mixed-Methods analysis
of bottom-up approaches in Brazilian public policy.}
\shortTitle{The Civil Society Organizations effect}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

\author{
  Manoel Galdino\\
  Bianca Vaz Mondo\\
  Juliana Mari Sakai\\
  Natália Paiva\thanks{
  Galdino: Universidade de São Paulo,
Brazil, \href{mailto:mgaldino@usp.br}{mgaldino@usp.br}.
  Mondo: Government Transparency Institute - Germany, \href{mailto:}{}.
  Sakai: Transparência Brasil - Brazil, \href{mailto:}{}.
  Paiva: Alandar Consulting - Germany, \href{mailto:}{}.
  We would like to thank all participants of project Obra Transparente,
  Vanda Medeiros and the team of Transparência Brasil who worked in the
  project or in evaluations of the project. The project was funded by an
  UNDEF grant, which we also thank.
}
}

\date{\today}
\pubMonth{11}
\pubYear{2023}
\pubVolume{}
\pubIssue{}
\JEL{}
\Keywords{accountability, Civil Society Organizations, mixed
methods, community-driven development, public policy.}

\begin{abstract}
This paper examines the effect of bottom-up accountability on public
service delivery. We differentiate between information-driven
interventions and the mobilization and monitoring efforts of organized
Civil Society Organizations (CSOs), and argue that the latter type of
interventions can can drive significant policy change. The study
evaluates the effectiveness of the Obra Transparente project by
Brazilian NGO Transparência Brasil, engaging 21 local CSOs in South and
Southeast Brazil. Our findings emphasize the importance of sustained,
coordinated efforts by socially embedded CSOS. These efforts, involving
direct engagement with municipal officials and ensuring that their
complains cannot be ignored lead to more substantial outcomes when
compared with citizens information-driven.
\end{abstract}

\maketitle

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Accountability is a crucial element in democratic systems, ensuring the
efficient delivery of public services (Besley and Ghatak 2003; Cameron
2004; O'donnell 1998). Initially defined as the process of holding
authorities responsible for their actions (Mulgan 2000; O'Loughlin
1990), the literature has since broadened to include various oversight
mechanisms for politicians and the bureaucracy. Some authors focus on
formal institutions (O'donnell 2003; Kenney 2003), known as horizontal
accountability, while others examine how politicians respond to voters,
the media, and civil society organizations (Bertelli and Van Ryzin 2020;
Relly 2012; Paul 1992), referred to as vertical accountability.

Recent studies have highlighted the role of ``bottom-up accountability''
(also known as social accountability, Fox, 2015), in which citizens
receive information about public policies and use it for social control.
The pioneering study `Power to the People' (Björkman and Svensson 2009),
which demonstrated the impact of local community engagement in
monitoring the provision of health services, inspired subsequent
literature that tested similar interventions in different contexts.

The mechanism would works like this: Constituents have firsthand
information about the outcomes of local policies, giving citizens
incentives to combat corruption that directly affects them. Since
policymakers are sensitive to social feedback from their own communities
(Serra 2012), bottom-up accountability could influence local public
policies

However, a series of recent studies have raised doubts about the
effectiveness of ``bottom-up accountability'' strategies based on
interventions aimed at increasing the monitoring capacity of ordinary
citizens regarding public policies (Freire, Galdino, and Mignozzetti
2020; Raffler, Posner, and Parkerson 2019; Fox 2015).

@@

The disappointing results have cast doubts on the effectiveness of the
bottom-up accountability strategy for improving the delivery of public
services, reducing corruption, and enhancing government efficiency. Some
authors (Fox 2015; Ankamah 2019) have suggested that social
accountability without ``teeth,'' defined as the government's capacity
to respond to citizens' voices, has limited impact.

While it has been consistently shown that tactical interventions have,
at best, a small, limited impact, it doesn't mean that voice without
``teeth'' is an ineffective strategy. The argument in this study is that
not all types of interventions that promote citizen's voices are the
same.

We argue that we need to differentiate between information-led tactical
interventions and active mobilization and monitoring by Civil Society
Organizations (CSOs). By their very nature, established CSOs are
embedded in society and politics and can have a substantial impact on
policy change (Gong and Xiao 2017; Joshi and Houtzager 2012).

The paper argues that a missing feature in most of the literature is
that these interventions should target CSOs instead of individual
citizens, and ensure that the intervention enhances the ability of the
CSOs to overcome their collective action problem (Mattoni and Odilla
2021; Odilla 2023).

The present paper provides evidence supporting this argument. We
reassess the evidence provided by Freire, Galdino, and Mignozzetti
(2020), considering a different intervention than the one they
evaluated. They provided experimental evidence that a mobile app (Tá de
Pé) made available to citizens did not improve the delivery of school
and nursery construction works. Here, we assess the efficacy of a sister
project, called ``Obra Transparente,'' also developed by the Brazilian
NGO Transparência Brasil. This project targeted 21 local Civil Society
Organizations in 21 municipalities in the South and Southeast regions of
Brazil, with the aim of improving access to education by addressing
mismanagement and delays in the construction of public schools and
nurseries. Both projects were developed in parallel, with similar goals
but achieved strikingly different results.

The argument in the present study is that not all types of interventions
that promote citizens' voice are the same. We argue that we need to
differentiate between information-led tactical interventions and active
mobilization and monitoring by Civil Society Organizations. By their
very nature, established CSOs are embedded in society and politics and,
as such, can have teeth of their own (Gong and Xiao 2017; Joshi and
Houtzager 2012). It may be only that we need mobilization by organized
civil society, in a structured way, for civil society voices to be heard
with ``teeth''. Some studies have suggested that a key feature of
successful change in public policies is the pressure of collectively
organized agents (Mattoni and Odilla 2021; Odilla 2023). Collective
action is a hard problem to be overcome and creating CSOs is one way to
achieve this. We argue in this paper that a missing features of the
tactical interventions assessed in most of the literature is that they
should target CSOs instead of individual citizens, and make sure that
the intervention increases the ability of the CSOs to attract and
overcome their collective action problem.

The ``Obra Transparente'' project consisted of developing a network of
local CSOs to share experiences and information on monitoring the
construction works, providing a technical support system for monitoring
the process of constructing public schools and nurseries, and training
on monitoring biddings, contracts, and constructions of public schools
and nurseries.

@@

\hypertarget{the-project-obra-transparente}{%
\subsection{The project Obra
Transparente}\label{the-project-obra-transparente}}

@@

\hypertarget{research-design-and-methods}{%
\subsection{Research Design and
Methods}\label{research-design-and-methods}}

We employ an integrated inference approach (Humphreys and Jacobs 2023)
to analyze qualitative data within the framework of causal queries. As
this approach is relatively innovative within qualitative research,
we'll provide a comprehensive explanation.

The methodology outlined by (Humphreys and Jacobs 2023) mandates the
creation of a causal model encoded in Directed Acyclic Graphs (DAGs), a
framework developed by Pearl (2009). Bayesian inference, recently
promoted in qualitative data analysis (Fairfield and Charman 2022),
serves as the foundation for deriving quantitative estimates for causal
queries constructed using the proposed DAG.

In our current study, we investigate the impact of Civil Society
Organization (CSO) monitoring on two aspects: the completion of
construction works and the rectification of issues and irregularities
found in construction facilities. Formally, the first research question
relates to a causal relationship between monitoring (M) and completion
(C), indicated by the DAG (M -\textgreater{} C). The second research
query delves into whether monitoring (M) causes the fixing of issues
(F), or the (DAG M -\textgreater{} F). To accommodate the qualitative
nature of our variables, we treat them as binary.

@@

This approach hinges on the notion that each case can belong to a
specific type, denoting different causal relationships. In our current
investigation, let's focus on the (M -\textgreater{} C) DAG. To clarify
what a type is, we'll employ the Potential Outcomes (PO) notation (Rubin
2005; Imbens and Rubin 2015). Within the PO framework, every case has
two potential outcomes: one when the treatment (in this case, CSO
monitoring, M) is absent and another when the treatment is present. In
simplified notation, if a city only completes construction works in the
absence of monitoring, we express this in PO notation as C(1) = 0. If
there is no monitoring (M = 0) and no completion, then we write C(0) =
0. Conversely, if there is monitoring (M = 1) and completion, we write
C(1) = 1, and C(0) = 1 if there is completion without monitoring.

We can categorize a city into one of four types based on what happens
when they do or do not receive the treatment (monitoring): Adverse (type
``a''): A city that completes construction only if it doesn't receive
the treatment (monitoring). Beneficial (type ``b''): A city that
completes construction only if there is monitoring. Chronic (type
``c''): Cities that never complete construction, regardless of
monitoring. Destined (type ``d''): Cities that always complete
construction, regardless of monitoring. The PO notation is instrumental
here, as we can rewrite the types based on the following notation: Type
``a'' is one in which the potential outcome is 0 if M = 1 and 1 if M =
0. Therefore, we express it as \(\theta_{10}\), where the first
subscript denotes the outcome when the treatment is 0, and the second
denotes the outcome when the treatment is 1. We employ analogous
notation for all other types: \(b = \theta_{01}\), \(c = \theta_{00}\),
and \(d = \theta_{11}\).

When analyzing a case study of what happened in a given municipality, we
can use the data to answer, for example, the following causal query:
what is the probability that the city if type ``b'', i.e, benefits from
the monitoring and would have not completed the construction work if not
for the monitoring of the local CSO?

The advantage of this approach is that all assumptions and restrictions
imposed to identify the parameters of interest are transparent. It also
allows the user to update posterior probabilities based on qualitative
data. One reason for this is that with Bayesian inference, we do not
need to rely on asymptotic theory to make inferences. Thus, there is no
need for a large-n study. Formally speaking, one can make causal queries
without any data, based only on prior probabilities for the parameters.
Obviously the answer to such causal queries will be vague enough to be
almost useless. But the point is that, once you have a prior and a
causal model, encoded in a DAG, you can answer causal queries. As one
collects data (any data, be it qualitative or quantitative), one can
make assumptions about the nature of the data (such as the Data
Generating Process), and then update the model and provide new answers
to the causal queries, which hopefully will be informative enough to be
useful for the task at hand. And even when the model or hypothesis is
not completely identified, it can be partially identified (Manski 2003),
which is still a useful answer to causal queries.

\hypertarget{data}{%
\subsection{Data}\label{data}}

@@

\hypertarget{case-studies}{%
\subsection{Case Studies}\label{case-studies}}

\hypertarget{delivery-of-the-construction-works}{%
\subsubsection{Delivery of the construction
works}\label{delivery-of-the-construction-works}}

@@

To put in perspective what was accomplished regarding the completion of
all four projects, by the end of the project, of all 135 construction
works planned for all 21 municipalities, only 25 were effectively
delivered, a 18.5\% delivery rate. In Taubaté, before the monitoring by
the local CSO, the delivery rate was zero. After that, the delivery rate
was 100\% and one of the construction work was a previously paralyzed
construction. According to a report by Transparência Brasil (Coelho,
Galdino, and Sakai 2021), out of 771 stalled construction projects that
were eventually completed, over 50\% of them took more than two years to
be delivered and 25\% took three years or more. As we can see, the odds
were against finishing the construction works.

The local CSO monitoring the construction work in Taubaté did one of the
best jobs among all 21 municipalities covered in the project. In fact,
it was mostly the work of a single volunteer with support from the local
OSB and Transparência Brazil throughout the project. She monitored four
ongoing construction works on a periodic basis, she was very
knowledgeable and insistent in demanding response from contractors and
the local government. To give a single example of the quality of the
social monitoring going on, in an email sent on 17th January of 2018 to
the local government, the local OSB volunteer wrote: ``Fazendinha
Daycare: a. construction log verified. There have been a few more
construction workers on the site since December 15, 2017, but not
enough, especially considering the extensive amount of work to be
completed by the new delivery date in April 2018. b. (\ldots). c.~Lack
of roof sealing, leading to water entering the classrooms (a situation
that already existed on November 14, 2017)\ldots{}'' . She was by far
the most knowledgeable and zealous volunteer of the whole project.

@@

From a causal perspective, we are interested in what is called
case-level causal attribution (Humphreys and Jacobs 2023), i.e, did the
monitoring cause the delivery of the construction works? In other words,
we are conditioning on the fact that M = 1 and C = 1, which constrains
the possible types for Taubaté to be \(\theta^C_{1,1}\) ou
\(\theta^C_{0,1}\). Assuming a simple model in which CSO monitoring
causes the completion of works (M -\textgreater{} C), and with the eight
within-case instances, we do the causal query. Our stimand of interest
is the probability that Taubaté is a type \(\theta^D_{10}\).

Assuming uniform priors, the probability that Taubaté is a type
\(\theta^D_{10}\), i.e, it is a city in which the treatment has a
positive effect is 65\%, with 95\% credibility interval around {[}23\%,
95\%{]}. As we can see, a wide interval, due to the low sample size, and
yet, a positive effect nonetheless with 95\% of certainty.
